---
title: "NFL Data"
author: "Hayden Rogers"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/hroge/OneDrive/Documents/datasets')
```

# Traversing and Cleaning the Data

```{r}
# Loading the ggplot2 and dplyr libraries
library(ggplot2)
library(dplyr)

# Reading the File
NFL <- read.csv("nfl_pass_rush_receive_raw_data.csv")

# Setting OT to a logical value
NFL$OT <- as.logical(NFL$OT)


# 3 Seasons
#     1 -  6330 -- Season 1 (unique games   1 - 266)
#  6331 - 12960 -- Season 2 (unique games 267 - 535)
# 12961 - 19973 -- Season 3 (unique games 536 - 820)

NFL_2019s <- NFL[c(1:6330), ]
NFL_2020s <- NFL[c(6331:12960), ]
NFL_2021s <- NFL[c(12961:19973), ]

# For getting individual's data
# Arodgers <- NFL[which(NFL$player_id == "RodgAa00"), ]

# Obtaining the unique game ids
game_ids <- unique(NFL$game_id)

# Getting the unique game ids for each season's games
game_ids_2019 <- unique(NFL_2019s$game_id)
game_ids_2020 <- unique(NFL_2020s$game_id)
game_ids_2021 <- unique(NFL_2021s$game_id)

# Filtering the NFL game data to each unique game's data
uniq_games <- NFL %>% 
  distinct(game_id, .keep_all = TRUE) %>%
  select(game_id, home_score, vis_score, home_team, vis_team, OT, Vegas_Favorite)

# Filtering for games that went to Overtime
OT_games <- filter(uniq_games, OT == TRUE)



```

# Questions

### How helpful is the home field advantage for underdogs?

For those who don’t know what a home field advantage is, Wikipedia describes the phenomenon as, [quote]. This can be seen in a number of ways, through the visiting team acclimating to a different climate or the home team gaining more support from increased numbers of fans.

```{r}

# Removing ties because there are a very low number of tied games
no_ties <- uniq_games %>% 
  filter(home_score - vis_score != 0) %>%
  mutate(home_under = Vegas_Favorite != home_team) %>%
  mutate(point_diff = home_score - vis_score) %>%
  mutate(home_won = (point_diff > 0)) %>%
  mutate(pos_diff = abs(point_diff)) %>%
  mutate(under_won = ((home_under & home_won) | (!home_under & !home_won)))


# Adding Labels for the Bar Graphs
no_ties$home_under <- factor(no_ties$home_under, 
                             levels = c(FALSE, TRUE),
                             labels = c("Visiting", "At Home"))

no_ties$under_won <- factor(no_ties$under_won, 
                             levels = c(FALSE, TRUE),
                             labels = c("Underdog Lost", "Underdog Won"))

# Underdog Win/Lose Stacked Bar Graph over Home/Visiting
ggplot(data = no_ties, mapping = aes(x = home_under, fill = under_won)) +
  geom_bar(position = "fill") +
  labs(title = "Does Home Field Advantage Help the Underdog?",
       x = "Underdog Location",
       y = "Proportion",
       fill = "Game Outcome")
```



```{r}
# Underdog Win/Lose Side-by-side Bar Graph over Home/Visiting
ggplot(data = no_ties, mapping = aes(x = home_under, fill = under_won)) +
  geom_bar(position = "dodge") +
  labs(title = "Does Home Field Advantage Help the Underdog?",
       x = "Underdog Location",
       y = "Number of Games",
       fill = "Game Outcome")

# Factorizing the difference in points to relevant values
no_ties$pos_diff <- cut(as.numeric(no_ties$pos_diff),
                             breaks = c(0, 7, 14, 21, 100),
                             labels = c("1-7 point difference", 
                                        "8-14 point difference", 
                                        "15-21 point difference", 
                                        "22+ point difference"))

# Adding Labels for 
no_ties$home_under <- factor(no_ties$home_under, 
                             levels = c("Visiting", "At Home"),
                             labels = c("Underdog Visiting", "Underdog at Home"))

# Point Difference Graph in Counts
ggplot(data = no_ties, mapping = aes(x = under_won, fill = as.factor(pos_diff))) +
  geom_bar() +                     # Unsure about adding a position 
  facet_wrap(home_under ~ .) +
  labs(title = "Does Home Field Advantage Help the Underdog?",
       fill = "Point Difference",
       y = "Number of Games",
       x = "Underdog Outcome")

# Point Difference Graph in Proportions
ggplot(data = no_ties, mapping = aes(x = under_won, fill = as.factor(pos_diff))) +
  geom_bar(position = "fill") +
  facet_wrap(home_under ~ .) +
  labs(title = "Does Home Field Advantage Help the Underdog?",
       fill = "Point Difference",
       y = "Proportion of Games",
       x = "Underdog Won")


```